Secure Composition of Robust and Optimising Compilers

### 2023-07-06 popl/main.tex
Text up to section 5 (excluding)

This section presents the key results of this paper that have been presented in prior work~\cite{kruse2022csc}.
The main results are the fact that secure compilers in the robust preservation framework~\cite{abate2019jour} compose.
...
Most notably, it looks at the sequential composition of secure compilers and shows a theorem that allows for modular proof engineering: compilers need not to robustly preserve the same property in order to be securely composed.
Moreover, it demonstrates that the order does not matter for composing compilers with the same source and target language.
Throughout this section, let $\src{L},\trg{L},$ and $\obj{L}$ be arbitrary programming lanugages.

\begin{definition}[Sequential Composition]\label{def:seqcomp}
  Let $\cc{\src{L}}{\trg{L}}$, $\cc{\trg{L}}{\obj{L}}$ be compilers from languages $\src{L}$ to $\trg{L}$ and $\trg{L}$ to $\obj{L}$, respectively.
  Their sequential composition is $\left(\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}\right)\left(\src{p}\right)=\obj{p}$.
\end{definition}

\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtp}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}$ and $\rtp{\cc{\trg{L}}{\obj{L}}}{\class[_{2}]}$.
  Then, $\rtp{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}$. \Coqed
\end{theorem}

A corollary of \Cref{thm:rtp} is that optimization passes that have the input and output language can be freely swapped:
\begin{corollary}[Sequential Composition of Secure Compilers]\label{corr:swappable}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{1}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{2}]}$.
  Then, $\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}\circ\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{1}]\cap\class[_{2}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}\circ\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{2}]\cap\class[_{1}]}$. \Coqed
\end{corollary}

>>>>

This section presents the key meta-theoretic results of this paper and extends on prior work~\cite{kruse2022csc}.
A main result is that secure compilers in the robust preservation framework~\cite{abate2019jour} compose.
This is not intuitive in the sense that there are countless examples in the security domain where composition does not work.\MK{cite}
For starters, the sequential composition of compilers $\cc{\src{L}}{\trg{L}}$ and $\cc{\trg{L}}{\obj{L}}$ is as expected, i.e., given an $\src{L}$ program $\src{p}$, obtain its compiled $\obj{L}$ program by plugging $\src{p}$ into $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$.
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtp}
  Given $\class[_{1}],\class[_{2}]$, \bul{$\rtp{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}$}, and \rul{$\rtp{\cc{\trg{L}}{\obj{L}}}{\class[_{2}]}$}, then \oul{$\rtp{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}$}. \Coqed
\end{theorem}

Since the composition of secure compilers is again a secure compiler, the theorem generalizes to a whole chain of $n$ secure compilers.
Pretending that the following compilers are secure, consider the compilation chain for $\src{TypeScript}$.
First, $\src{TypeScript}$ programs are translated to $\trg{JavaScript}$ which, e.g., V8\MK{cit} eventually compiles in parts to $\irl{Ignition BC}$.
Were all these compilation steps robustly secure with respect to \gls{ms}, the resulting $\irl{Ignition BC}$ code would be \gls{ms} regardless of the context the binary runs in.

Notably, real-world compilation chains also perform a series of passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimize the code or enforce a certain property.
Both examples can be seen in the wild, there are countless instances of them.\MK{cit}
Consider the compilation passes of LLVM\MK{cit} whose source-code optimisations are industry--competitive.
In LLVM, the pass manager can be used to schedule passes in a way where the optimisations yield the best results.
Take \gls{cf} and \gls{dce}.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration.
However, this costs additional time until compilation is done, which can be a valuable resource\MK{cit}.
Hence, compiler engineers typically try to find an order of optimizations that yields well-optimized programs as fast as possible.
The following corollary justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

\begin{corollary}[Sequential Composition of Secure Compilers]\label{corr:swappable}
  Given $\class[_{1}],\class[_{2}]$, \bul{$\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{1}]}$, and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{2}]}$}, then \oul{$\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}\circ\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{1}]\cap\class[_{2}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}\circ\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{2}]\cap\class[_{1}]}$}. \Coqed
\end{corollary}

Besides the sequential composition, there are two other compositions that are more at the language- than the compiler--level.
Nevertheless, consider $\trg{JVM BC}$, which is a popular target for programming language designers due to its high performance and relevance in industry.
As an example, $\src{Java}$ and $\obj{Kotlin}$ compile to $\trg{JVM BC}$.
Technically speaking, they both lower to class files and $\obj{Kotlin}$ objects are considered to be the same as $\src{Java}$ objects at that point.
A compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$ code translating to the same target language, or rather, intermediate representation, performs a kind of {\em upper} composition.
Given a secure compiler from $\src{Java}$ to $\trg{JVM BC}$ and one from $\obj{Kotlin}$ to $\trg{JVM BC}$, it is easy to envision that one can engineer a combined compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$.
A practical example of this, which is, however, insecure, is {\tt Android Studio}\MK{cit}.
The proof of \Cref{thm:rtp} can be easily adapted to also works for this {\em upper} composition.
The corresponding theorem is not listed, since there is no general definition of this kind of composition.

Likewise, the {\em lower} composition is concerned about compilers that accept the same source but yield different target languages and the same theoretical results apply.
However, especially for this kind of composition, it is rather difficult to provide a general definition.
Consider two compilers both accepting $\src{LLVM IR}$ and one of them emits $\trg{x86\_64}$, while the other emits $\trg{ARMv8}$.
It is intuitive that they are in some sense composed in the LLVM framework, but the decision of when to use one over the other is inherently {\em unobservable} to the formalization effort of this kind of composition.
Either the user of LLVM provides an explicit flag that instructs the overall compiler to emit $\trg{x86\_64}$ or the framework itself detects the target platform via heuristics.
%Either way, the theoretical results suggest that it is possible to always find a ,,most--general'', secure compiler, given two secure compilers, that robustly preserves the least--upper bound of the classes involved in their compilation process.
In general, there are two options to deal with these nuances in a concrete, practical setting:
Given secure compilers, either (i) view the whole compilation chain involving them as the upper/sequential/lower compositions and conclude the robust preservation of the intersection of all involved classes, or (ii) specify meta-parameters that, e.g., specify the target language, and thus reduce the compilation chain to a nested sequential composition of the involved compilers.
The first approach yields a more imprecise statement about the whole compilation chain while the second approach is more precise, but also less flexible.

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimizations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimized towards the righthandside of the figure.}\label{fig:cfdceex}
\end{figure}
